# Exploratory data analysis

How can we explore transcript expression from long read data...

Let's load some data...

```{r}
library(here)
url <- "https://github.com/gandallab/Dev_Brain_IsoSeq/raw/main/data/cp_vz_0.75_min_7_recovery_talon_abundance_filtered.tsv.gz"
filename <- basename(url)
path <- here("data",filename)
if (!file.exists(path)) {
  download.file(url, path)
}
```

```{r}
library(readr)
raw_abundance_table <- read_delim(path)
dim(raw_abundance_table)
colnames(raw_abundance_table)
raw_abundance_table[1:5,1:15]
```

```{r}
library(dplyr)
library(tidyr)
raw_abundance_table |>
  select(contains(c("VZ","CP"))) |>
  summarize(across(everything(), sum))
```

...maybe now move some of the processing code from `testing.qmd`... we
only need to do this once, and then we could save an intermediate
object and re-load it for other chapters, so we aren't always starting
a chapter with the bookkeeping stuff. maybe we could have an earlier
chapter just on loading/munging data from a flat file into a
Bioconductor object. 
